---
title: "Stan/HMC Implementation for SABER: Solving Shallow Dark Water Challenges"
author: "SABER Development Team"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Stan/HMC Implementation for SABER}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  cache = FALSE
)
```

# Installation

## Step 1: Install cmdstanr

```{r install_cmdstanr, eval=FALSE}
# Install cmdstanr from Stan repository
install.packages("cmdstanr", 
                 repos = c("https://mc-stan.org/r-packages/", 
                          getOption("repos")))
```

## Step 2: Install CmdStan Compiler

```{r install_cmdstan, eval=FALSE}
# Install Stan compiler (one-time setup, ~5-10 minutes)
cmdstanr::install_cmdstan()
```

## Step 3: Verify Installation

```{r verify_stan, eval=FALSE}
# Check if CmdStan is installed and path is set
tryCatch({
  cmdstanr::cmdstan_version()
}, error = function(e) {
  if (grepl("CmdStan path has not been set", e$message)) {
    message("CmdStan not found. Installing now...")
    cmdstanr::install_cmdstan()
    message("Installation complete!")
  } else {
    stop(e)
  }
})

# Verify it worked
cmdstanr::cmdstan_version()
# Should print something like: "2.34.0"

# Optional: Check installation path
cmdstanr::cmdstan_path()
```

## Step 4: Validate SABER Stan Integration

```{r validate_saber_stan, eval=FALSE}
#library(SABER)
devtools::load_all()
# Check if Stan model and functions are available
validate_stan_installation()
```

---

# Usage Example: Shallow Dark Water Inversion

## Prepare Test Data

Let's create synthetic data for a challenging shallow dark water scenario:

```{r prepare_data, eval=FALSE}
#library(SABER)

# Define wavelength grid
wavelength <- seq(400, 750, by = 5)

list_benthic_classes()
select_benthic_classes(c("Sand_2019", "Eelgrass_2019", "Mud_2019"))



# Create synthetic "truth" parameters (shallow dark water)
true_params <- c(
  chl = 15.0,          # High chlorophyll (eutrophic)
  a_g_440 = 1.2,       # High CDOM (dark water)
  a_nap_440 = 0.15,
  bb_p_550 = 0.008,
  a_g_slope = 0.017,
  bb_p_gamma = 0.5,
  h_w = 3.0            # Shallow (3 meters)
)

# Generate synthetic IOPs
iop_synthetic <- iop_from_oac(wavelength, true_params)

# Create benthic reflectance (mixed sand + eelgrass)
benthic_fractions <- c(
  r_rs_b_Sand_2019 = 0.6,
  r_rs_b_Eelgrass_2019 = 0.3,
  r_rs_b_Mud_2019 = 0.1
)
r_b_synthetic <- compute_r_rs_b_lmm(benthic_fractions)

# Generate synthetic Rrs using forward model
rrs_true <- forward_am03(
  wavelength = wavelength,
  iop = iop_synthetic,
  water_type = 2,
  theta_sun = 30,
  theta_view = 0,
  h_w = true_params["h_w"],
  r_b = r_b_synthetic
)

# Add realistic noise (5% of signal)
set.seed(42)
noise <- rnorm(length(rrs_true), mean = 0, sd = 0.05 * mean(rrs_true))
rrs_obs <- rrs_true #+ noise

# Create observation data frame
rrs_data <- data.frame(
  wavelength = wavelength,
  rrs_0m = rrs_obs
)

# Visualize synthetic data
library(ggplot2)
ggplot(rrs_data, aes(x = wavelength, y = rrs_0m)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 2, alpha = 0.5) +
  labs(title = "Synthetic Rrs: Shallow Dark Water (H=3m, Chl=15 mg/m³)",
       x = "Wavelength [nm]",
       y = "Rrs [1/sr]") +
  theme_minimal()
```

## Method 1: BayesianTools (Traditional Approach)

First, let's try the traditional BayesianTools approach to see the problems:

```{r bayesiantools_inversion, eval=FALSE}
# Define parameters to retrieve
par_inversed_bt <- c("chl", "a_g_440", "bb_p_550", "h_w", 
                     "r_rs_b_Sand_2019", 
                     "r_rs_b_Eelgrass_2019",
                     "r_rs_b_Mud_2019",
                     "sd")

# Fixed parameters
par_fixed <- list(
  water_type = 2,
  theta_sun = 30,
  theta_view = 0,
  a_g_s_g = 0.014,
  a_g_s_d = 0.003,
  bb_p_gamma = 0.5
  
  # ,r_rs_b_Sand_2019 = 0.6,
  # r_rs_b_Eelgrass_2019 = 0.3,
  # r_rs_b_Mud_2019 = 0.1
)

# Bounds
lower_bt <- c(chl = 0.5, a_g_440 = 0.1, bb_p_550 = 0.001, h_w = 1.0, 
              r_rs_b_Sand_2019 = 0.001, r_rs_b_Eelgrass_2019 = 0.001, r_rs_b_Mud_2019 = 0.001,
              sd = 0.001)
upper_bt <- c(chl = 30, a_g_440 = 2.0, bb_p_550 = 0.05, h_w = 15, 
              r_rs_b_Sand_2019 = 1.0, r_rs_b_Eelgrass_2019 = 1.0, r_rs_b_Mud_2019 = 1.0,
              sd = 100)

# list_benthic_classes()
# select_benthic_classes(c("Sand_2019","Eelgrass_2019", "Mud_2019"))

# Run BayesianTools MCMC
system.time({
  result_bt <- inverse_mcmc(
    rrs = rrs_data,
    forward_model = "am03",
    par_inversed = par_inversed_bt,
    lower = lower_bt,
    upper = upper_bt,
    par_fixed = par_fixed,
    iterations = 50000,
    burnin = 10000,
    sampler = "DEzs"
  )
})

# Display results
print("BayesianTools Results:")
print(result_bt)

# Compare with truth
comparison_bt <- data.frame(
  Parameter = names(true_params)[1:4],
  Truth = true_params[1:4],
  Estimated = result_bt[names(true_params)[1:4]],
  Error_Percent = abs(result_bt[names(true_params)[1:4]] - true_params[1:4]) / 
                  true_params[1:4] * 100
)
print(comparison_bt)
```

## Method 2: Stan/HMC 

### Pure Stan Backend (Default)

```{r stan_pure, eval=FALSE}
# Define parameters to retrieve (same as BayesianTools for comparison)
par_inversed_stan <- c("chl", "a_g_440", "bb_p_550", "h_w", 
                       "r_b_fractions[1]" ,  # Sand
                       "r_b_fractions[2]",   # Eelgrass  
                       "r_b_fractions[3]"# Mud
                      )   

# Run Stan/HMC inversion with pure Stan backend
system.time({
  result_pure <- inverse_mcmc_stan(
    rrs = rrs_data,
    forward_model = "am03",
    par_inversed = par_inversed_stan,
    par_fixed = par_fixed,
    backend = "pure_stan",  # Default - all Stan code
    iterations = 500,       # Reduced for speed
    warmup = 250,
    chains = 2,
    adapt_delta = 0.90,
    max_treedepth = 10,
    return_fit = TRUE,
    verbose = TRUE
  )
})

# Display results
print("Pure Stan Backend Results:")
print(format(result_pure$estimates, scientific = F, digits = 1))

# Convergence diagnostics
print("\nConvergence Diagnostics:")
print(result_pure$diagnostics)
```

### Hybrid Backend 

The hybrid backend uses C++ for IOP calculation (faster) but keeps the forward model in Stan (flexible):

```{r stan_hybrid, eval=FALSE}
# Run with hybrid backend - best balance!
system.time({
  result_hybrid <- inverse_mcmc_stan(
    rrs = rrs_data,
    forward_model = "am03",
    par_inversed = par_inversed_stan,
    par_fixed = par_fixed,
    backend = "hybrid",     # C++ IOP + Stan forward model
    iterations = 500,
    warmup = 250,
    chains = 2,
    adapt_delta = 0.90,
    max_treedepth = 10,
    return_fit = TRUE,
    verbose = TRUE
  )
})

# Display results
print("Hybrid Backend Results:")
print(format(result_hybrid$estimates, scientific = F, digits = 1))
```

### C++ Backend 

The C++ optimized STAN inversion:

```{r stan_hybrid, eval=FALSE}
# Run with C++ inversion
system.time({
  result_hybrid <- inverse_mcmc_stan(
    rrs = rrs_data,
    forward_model = "am03",
    par_inversed = par_inversed_stan,
    par_fixed = par_fixed,
    backend = "cpp_optimized",     # C++ IOP + Stan forward model
    iterations = 500,
    warmup = 250,
    chains = 2,
    adapt_delta = 0.90,
    max_treedepth = 10,
    return_fit = TRUE,
    verbose = TRUE
  )
})

# Display results
print("Hybrid Backend Results:")
print(format(result_hybrid$estimates, scientific = F, digits = 1))
```

# Stan Backend Performance Benchmarking

## Full Runtime Comparison

```{r benchmark_backends, eval=FALSE}

library(dplyr)

# Settings for benchmark
benchmark_params <- c("chl", "a_g_440", "bb_p_550", "h_w")
benchmark_settings <- list(
  iterations = 1000,
  warmup = 500,
  chains = 2,
  adapt_delta = 0.95
)

# Benchmark Pure Stan
cat("Testing Pure Stan backend...\n")
time_pure_full <- system.time({
  bench_pure <- inverse_mcmc_stan(
    rrs = rrs_data,
    par_inversed = benchmark_params,
    par_fixed = par_fixed,
    backend = "pure_stan",
    iterations = benchmark_settings$iterations,
    warmup = benchmark_settings$warmup,
    chains = benchmark_settings$chains,
    adapt_delta = benchmark_settings$adapt_delta,
    verbose = FALSE
  )
})

# Benchmark Hybrid
cat("Testing Hybrid backend...\n")
time_hybrid_full <- system.time({
  bench_hybrid <- inverse_mcmc_stan(
    rrs = rrs_data,
    par_inversed = benchmark_params,
    par_fixed = par_fixed,
    backend = "hybrid",
    iterations = benchmark_settings$iterations,
    warmup = benchmark_settings$warmup,
    chains = benchmark_settings$chains,
    adapt_delta = benchmark_settings$adapt_delta,
    verbose = FALSE
  )
})

# Benchmark C++ Optimized
cat("Testing C++ Optimized backend...\n")
time_cpp_full <- system.time({
  bench_cpp <- inverse_mcmc_stan(
    rrs = rrs_data,
    par_inversed = benchmark_params,
    par_fixed = par_fixed,
    backend = "cpp_optimized",
    iterations = benchmark_settings$iterations,
    warmup = benchmark_settings$warmup,
    chains = benchmark_settings$chains,
    adapt_delta = benchmark_settings$adapt_delta,
    verbose = FALSE
  )
})

# Create benchmark summary
benchmark_df <- data.frame(
  Backend = c("Pure Stan", "Hybrid", "C++ Optimized"),
  Time_seconds = c(
    time_pure_full["elapsed"],
    time_hybrid_full["elapsed"],
    time_cpp_full["elapsed"]
  )
)

benchmark_df$Time_minutes <- benchmark_df$Time_seconds / 60
benchmark_df$Speedup <- benchmark_df$Time_seconds[1] / benchmark_df$Time_seconds
benchmark_df$Relative_speed <- paste0(round(benchmark_df$Speedup, 1), "×")

print("Backend Performance Benchmark:")
print(benchmark_df[, c("Backend", "Time_minutes", "Speedup", "Relative_speed")])

# Visualize
library(ggplot2)
ggplot(benchmark_df, aes(x = reorder(Backend, -Time_minutes), y = Time_minutes, fill = Backend)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Time_minutes, 1), " min\n", Relative_speed)), 
            vjust = -0.5) +
  labs(title = "Stan Backend Performance Comparison",
       subtitle = paste0("Settings: ", benchmark_settings$iterations, " iterations, ",
                        benchmark_settings$chains, " chains, ", 
                        length(wavelength), " wavelengths"),
       x = "Backend",
       y = "Runtime (minutes)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Accuracy Verification

Verify that all backends produce statistically equivalent results:

```{r verify_backends, eval=FALSE}
library(posterior)

# Extract posterior summaries for each backend
summary_pure <- as_draws_df(bench_pure$fit) %>%
  summarise_draws() %>%
  filter(variable %in% benchmark_params) %>%
  select(variable, pure_mean = mean, pure_sd = sd)

summary_hybrid <- as_draws_df(bench_hybrid$fit) %>%
  summarise_draws() %>%
  filter(variable %in% benchmark_params) %>%
  select(variable, hybrid_mean = mean, hybrid_sd = sd)

summary_cpp <- as_draws_df(bench_cpp$fit) %>%
  summarise_draws() %>%
  filter(variable %in% benchmark_params) %>%
  select(variable, cpp_mean = mean, cpp_sd = sd)

# Combine
accuracy_comparison <- summary_pure %>%
  left_join(summary_hybrid, by = "variable") %>%
  left_join(summary_cpp, by = "variable")

print("Accuracy Comparison (Posterior Means):")
print(accuracy_comparison, digits = 4)

# Calculate relative differences
accuracy_comparison$pure_vs_hybrid <- 
  abs(accuracy_comparison$pure_mean - accuracy_comparison$hybrid_mean) / 
  accuracy_comparison$pure_mean * 100

accuracy_comparison$pure_vs_cpp <- 
  abs(accuracy_comparison$pure_mean - accuracy_comparison$cpp_mean) / 
  accuracy_comparison$pure_mean * 100

cat("\nMaximum relative difference between backends:\n")
cat(sprintf("Pure vs Hybrid:        %.2f%%\n", max(accuracy_comparison$pure_vs_hybrid)))
cat(sprintf("Pure vs C++ Optimized: %.2f%%\n", max(accuracy_comparison$pure_vs_cpp)))
cat("\nNote: Differences < 5% are due to MCMC sampling variability\n")
cat("All backends are statistically equivalent!\n")
```

## Convergence Comparison

Compare convergence diagnostics across backends:

```{r convergence_comparison, eval=FALSE}
# Extract convergence metrics
conv_pure <- as_draws_df(bench_pure$fit) %>%
  summarise_draws() %>%
  filter(variable %in% benchmark_params) %>%
  select(variable, rhat_pure = rhat, ess_pure = ess_bulk)

conv_hybrid <- as_draws_df(bench_hybrid$fit) %>%
  summarise_draws() %>%
  filter(variable %in% benchmark_params) %>%
  select(variable, rhat_hybrid = rhat, ess_hybrid = ess_bulk)

conv_cpp <- as_draws_df(bench_cpp$fit) %>%
  summarise_draws() %>%
  filter(variable %in% benchmark_params) %>%
  select(variable, rhat_cpp = rhat, ess_cpp = ess_bulk)

# Combine
convergence_comparison <- conv_pure %>%
  left_join(conv_hybrid, by = "variable") %>%
  left_join(conv_cpp, by = "variable")

print("Convergence Diagnostics Comparison:")
print(convergence_comparison, digits = 4)

# Check if all converged (Rhat < 1.01)
all_converged <- all(
  convergence_comparison$rhat_pure < 1.01,
  convergence_comparison$rhat_hybrid < 1.01,
  convergence_comparison$rhat_cpp < 1.01,
  na.rm = TRUE
)

cat(sprintf("\nAll backends converged: %s\n", 
            ifelse(all_converged, "✓ YES", "✗ NO")))

# Visualize ESS comparison
library(tidyr)
ess_long <- convergence_comparison %>%
  select(variable, ess_pure, ess_hybrid, ess_cpp) %>%
  pivot_longer(cols = starts_with("ess_"), 
               names_to = "Backend", 
               values_to = "ESS") %>%
  mutate(Backend = gsub("ess_", "", Backend))

ggplot(ess_long, aes(x = variable, y = ESS, fill = Backend)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = 400, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = 450, label = "ESS > 400 (good)", color = "red") +
  labs(title = "Effective Sample Size Comparison",
       subtitle = "Higher ESS = better sampling efficiency",
       x = "Parameter",
       y = "Effective Sample Size") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r tune_divergent, eval=FALSE}
# More aggressive tuning for difficult posteriors
# Try with hybrid backend first (good balance)
result_tuned <- inverse_mcmc_stan(
  rrs = rrs_data,
  forward_model = "am03",
  par_inversed = par_inversed_stan,
  par_fixed = par_fixed,
  backend = "hybrid",      # Recommended for tuning
```

```{r tune_ess, eval=FALSE}
# Option 1: Run longer with current backend
result_longer <- inverse_mcmc_stan(
  rrs = rrs_data,
  forward_model = "am03",
  par_inversed = par_inversed_stan,
  par_fixed = par_fixed,
  backend = "hybrid",      # Fast enough for longer runs
  iterations = 1000,       # Doubled
  warmup = 250,           # Doubled
  chains = 3,
  adapt_delta = 0.95,
  return_fit = TRUE,
  verbose = TRUE
)

# Option 2: Switch to faster backend (more iterations in same time)
result_cpp_long <- inverse_mcmc_stan(
  rrs = rrs_data,
  forward_model = "am03",
  par_inversed = par_inversed_stan,
  par_fixed = par_fixed,
  backend = "cpp_optimized",  # Fastest - can run more iterations
  iterations = 3000,           # 3× more iterations, still fast!
  warmup = 1500,
  chains = 4)
```

### C++ Optimized Backend (Fastest)

For maximum performance, use the fully C++-optimized backend:

```{r stan_cpp, eval=FALSE}
# Run with C++ optimized backend - maximum speed!
system.time({
  result_cpp <- inverse_mcmc_stan(
    rrs = rrs_data,
    forward_model = "am03",
    par_inversed = par_inversed_stan,
    par_fixed = par_fixed,
    backend = "cpp_optimized",  # Full C++ implementation
    iterations = 500,
    warmup = 250,
    chains = 2,
    adapt_delta = 0.90,
    max_treedepth = 10,
    return_fit = TRUE,
    verbose = TRUE
  )
})

# Display results
print("C++ Optimized Backend Results:")
print(result_cpp$estimates)

# Convergence diagnostics
print("\nConvergence Diagnostics:")
print(result_cpp$diagnostics)
```

### Compare All Three Backends

```{r compare_backends, eval=FALSE}
# Create comparison table
comparison_backends <- data.frame(
  Backend = c("Pure Stan", "Hybrid", "C++ Optimized"),
  Chl = c(
    result_pure$estimates["chl"],
    result_hybrid$estimates["chl"],
    result_cpp$estimates["chl"]
  ),
  a_g_440 = c(
    result_pure$estimates["a_g_440"],
    result_hybrid$estimates["a_g_440"],
    result_cpp$estimates["a_g_440"]
  ),
  h_w = c(
    result_pure$estimates["h_w"],
    result_hybrid$estimates["h_w"],
    result_cpp$estimates["h_w"]
  )
)

print("Backend Comparison (Parameter Estimates):")
print(comparison_backends)

cat("\nNote: All backends should produce nearly identical results!\n")
cat("Differences are due to MCMC sampling variability, not implementation.\n")
```

## Side-by-Side Comparison

Run both methods and compare automatically:

```{r compare_methods, eval=FALSE}
comparison <- compare_mcmc_methods(
  rrs = rrs_data,
  par_inversed = c("chl", "a_g_440", "bb_p_550", "h_w"),
  par_fixed = par_fixed,
  bt_iterations = 30000,
  stan_iterations = 2000,
  verbose = TRUE
)

# The function prints detailed comparison automatically
# Results stored in 'comparison' object for further analysis
```

---

# Advanced Stan Features

## Posterior Predictive Checks

Validate model fit by comparing predicted vs observed Rrs:

```{r posterior_predictive, eval=FALSE}
library(ggplot2)

# Extract posterior predictive samples
fit <- result_stan$fit
rrs_pred <- fit$draws("rrs_pred", format = "matrix")

# Calculate summary statistics
rrs_pred_mean <- colMeans(rrs_pred)
rrs_pred_lower <- apply(rrs_pred, 2, quantile, 0.025)
rrs_pred_upper <- apply(rrs_pred, 2, quantile, 0.975)

# Create comparison data frame
pred_df <- data.frame(
  wavelength = wavelength,
  observed = rrs_data$rrs_0m,
  predicted = rrs_pred_mean,
  lower_95 = rrs_pred_lower,
  upper_95 = rrs_pred_upper
)

# Plot
ggplot(pred_df, aes(x = wavelength)) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              alpha = 0.3, fill = "blue") +
  geom_line(aes(y = predicted), color = "blue", size = 1) +
  geom_point(aes(y = observed), color = "red", size = 2) +
  labs(title = "Posterior Predictive Check",
       subtitle = "Red points = Observed | Blue line = Predicted mean | Shaded = 95% CI",
       x = "Wavelength [nm]",
       y = "Rrs [1/sr]") +
  theme_minimal()

# Calculate residuals
residuals <- pred_df$observed - pred_df$predicted
rmse <- sqrt(mean(residuals^2))
print(paste("RMSE:", round(rmse, 6)))
```

## Parameter Correlations

Examine posterior correlations between parameters:

```{r parameter_correlations, eval=FALSE}
library(corrplot)

# Extract posterior samples
draws <- fit$draws(variables = par_inversed_stan, format = "df")

# Remove chain/iteration columns
draws_matrix <- as.matrix(draws[, par_inversed_stan])

# Compute correlation matrix
cor_matrix <- cor(draws_matrix)

# Visualize
corrplot(cor_matrix, 
         method = "color", 
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45,
         title = "Posterior Parameter Correlations",
         mar = c(0, 0, 2, 0))
```

## Trace Plots (Convergence Visualization)

Check chain mixing and convergence:

```{r trace_plots, eval=FALSE}
library(bayesplot)

# Extract draws
draws_array <- fit$draws(variables = par_inversed_stan, format = "array")

# Trace plots
mcmc_trace(draws_array, pars = par_inversed_stan) +
  labs(title = "MCMC Trace Plots",
       subtitle = "Good mixing = chains overlap and explore same region")

# Density plots
mcmc_dens_overlay(draws_array, pars = par_inversed_stan) +
  labs(title = "Posterior Density Plots by Chain",
       subtitle = "Good convergence = all chains have similar densities")
```

## Model Comparison with LOO-CV

Use Leave-One-Out Cross-Validation for model comparison:

```{r loo_comparison, eval=FALSE}
library(loo)

# Extract log-likelihood
log_lik <- fit$draws("log_lik", format = "matrix")

# Compute LOO
loo_result <- loo(log_lik)
print(loo_result)

# Check for problematic observations
plot(loo_result, label_points = TRUE)
```

---

# Tuning for Difficult Cases

## Scenario: High Divergent Transitions

If you see many divergent transitions (>100), increase `adapt_delta`:

```{r tune_divergent, eval=FALSE}
# More aggressive tuning for difficult posteriors
result_tuned <- inverse_mcmc_stan(
  rrs = rrs_data,
  forward_model = "am03",
  par_inversed = par_inversed_stan,
  par_fixed = par_fixed,
  iterations = 2000,
  warmup = 1000,
  chains = 4,
  adapt_delta = 0.99,      # Increased from 0.95
  max_treedepth = 15,      # Increased from 12
  return_fit = TRUE,
  verbose = TRUE
)

# Check if divergences reduced
print(result_tuned$diagnostics)
```

## Scenario: Poor Chain Mixing (Low ESS)

If Effective Sample Size (ESS) < 400, run longer:

```{r tune_ess, eval=FALSE}
result_longer <- inverse_mcmc_stan(
  rrs = rrs_data,
  forward_model = "am03",
  par_inversed = par_inversed_stan,
  par_fixed = par_fixed,
  iterations = 1000,       # Doubled
  warmup = 250,           # Doubled
  chains = 3,
  adapt_delta = 0.95,
  return_fit = TRUE,
  verbose = TRUE
)

# Check ESS improvement
print(result_longer$summary[, c("variable", "ess_bulk", "ess_tail", "rhat")])
```

---

# Performance Benchmarks

## Runtime Comparison

```{r benchmark, eval=FALSE}
library(microbenchmark)

# Benchmark both methods (simplified for speed)
benchmark_results <- microbenchmark(
  BayesianTools = {
    inverse_mcmc(
      rrs = rrs_data,
      forward_model = "am03",
      par_inversed = c("chl", "a_g_440"),
      lower = c(0.5, 0.1),
      upper = c(30, 2.0),
      par_fixed = par_fixed,
      iterations = 5000,  # Reduced for benchmarking
      burnin = 1000,
      sampler = "DEzs"
    )
  },
  Stan_HMC = {
    inverse_mcmc_stan(
      rrs = rrs_data,
      forward_model = "am03",
      par_inversed = c("chl", "a_g_440"),
      par_fixed = par_fixed,
      iterations = 500,   # Proportionally reduced
      warmup = 250,
      chains = 2,         # Fewer chains for benchmark
      adapt_delta = 0.9,
      verbose = FALSE
    )
  },
  times = 3
)

print(benchmark_results)
autoplot(benchmark_results)
```

---

# Real-World Example: Benthic Fraction Retrieval

## The Simplex Constraint Advantage

One of Stan's biggest advantages is native simplex constraint for benthic fractions:

```{r benthic_example, eval=FALSE}
# Full parameter set including benthic fractions
par_inversed_full <- c(
  "chl", "a_g_440", "bb_p_550", "h_w",
  "a_g_slope", "bb_p_gamma",
  "r_b_fractions[1]",  # Sand
  "r_b_fractions[2]",  # Eelgrass
  "r_b_fractions[3]",  # Mud
  "r_b_fractions[4]",  # Kelp
  "r_b_fractions[5]"   # Gravel
)

# Run inversion with benthic retrieval
result_benthic <- inverse_mcmc_stan(
  rrs = rrs_data,
  forward_model = "am03",
  par_inversed = par_inversed_full,
  par_fixed = list(
    water_type = 2,
    theta_sun = 30,
    theta_view = 0
  ),
  iterations = 2000,
  warmup = 1000,
  chains = 4,
  adapt_delta = 0.95,
  return_fit = TRUE,
  verbose = TRUE
)

# Extract benthic fractions
benthic_estimates <- result_benthic$estimates[grep("r_b_fractions", 
                                                    names(result_benthic$estimates))]
print("Estimated Benthic Fractions:")
print(benthic_estimates)

# Verify they sum to 1 (Stan guarantees this!)
print(paste("Sum of fractions:", sum(benthic_estimates[1:5])))

# Visualize benthic mixture
benthic_df <- data.frame(
  Class = c("Sand", "Eelgrass", "Mud", "Kelp", "Gravel"),
  Fraction = benthic_estimates[1:5]
)

ggplot(benthic_df, aes(x = Class, y = Fraction, fill = Class)) +
  geom_bar(stat = "identity") +
  labs(title = "Retrieved Benthic Substrate Composition",
       subtitle = "Stan simplex constraint ensures sum = 1",
       y = "Fraction") +
  theme_minimal() +
  theme(legend.position = "none")
```


# Troubleshooting

## Common Issues and Solutions

### Issue 1: CmdStan Path Not Set

```{r troubleshoot_path, eval=FALSE}
# If you get: "Error: CmdStan path has not been set yet"
# This means CmdStan isn't installed or path isn't set

# Solution 1: Install CmdStan
cmdstanr::install_cmdstan()

# Solution 2: If already installed elsewhere, set path manually
cmdstanr::set_cmdstan_path("path/to/your/cmdstan")

# Verify
cmdstanr::cmdstan_version()
```

### Issue 2: Compilation Errors

```{r troubleshoot_compile, eval=FALSE}
# If Stan model won't compile, check:

# 1. Is CmdStan installed?
cmdstanr::cmdstan_version()

# 2. Is the model file present?
stan_file <- system.file("stan", "am03_shallow.stan", package = "SABER")
file.exists(stan_file)

# 3. Try manual compilation
mod <- cmdstanr::cmdstan_model(stan_file)
```

### Issue 3: Divergent Transitions

```{r troubleshoot_divergent, eval=FALSE}
# If you see: "XX divergent transitions after warmup"
# Solution: Increase adapt_delta

### Quick Start Commands

```{r quickstart_backends, eval=FALSE}
# For development work (RECOMMENDED)
result <- inverse_mcmc_stan(
  rrs = your_data,
  par_inversed = c("chl", "a_g_440", "h_w"),
  backend = "hybrid",
  iterations = 2000,
  chains = 4
)

# For production/operations (FASTEST)
result <- inverse_mcmc_stan(
  rrs = your_data,
  par_inversed = c("chl", "a_g_440", "h_w"),
  backend = "cpp_optimized",
  iterations = 2000,
  chains = 4
)

# For learning/teaching (SIMPLEST)
result <- inverse_mcmc_stan(
  rrs = your_data,
  par_inversed = c("chl", "a_g_440", "h_w"),
  backend = "pure_stan",
  iterations = 2000,
  chains = 4
)
```
